---
title: "NYPD Police Shooting (Historic) Data Analysis"
author: "Data Science Student"
date: "8/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
****
```{r,echo=TRUE}
# Loading libraries
library(tidyr)
library(ggplot2)
library(stringr)
library(lubridate)
library(dplyr)
```
****
### Reading Session Information
```{r, echo=TRUE}
sessionInfo()
```
****
## **NYPD Police Shooting (Historic) Data Analysis**
### Reading the URL and data
Reading the URL for NYPD Police shooting (Historic) data file download
```{r,echo=TRUE}
url_link <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nyc_ps_data <- read.csv(url_link)
# Sorting the data by INCIDENT_KEY
nyc_ps_data <- arrange(nyc_ps_data, INCIDENT_KEY)
```
### Removing unwanted data columns
```{r,echo=TRUE}
nyc_ps_data <- nyc_ps_data %>% select(-c("X_COORD_CD", "Y_COORD_CD", "Latitude", "Longitude", "Lon_Lat"))

# Omitting 3 rows with Perp age group not representing a proper range
nyc_ps_data <- nyc_ps_data %>% filter((PERP_AGE_GROUP != 224) & (PERP_AGE_GROUP != 940) & (PERP_AGE_GROUP != 1020))

str(nyc_ps_data)
```
### Formatting Date and time columns
```{r,echo=TRUE}
# Date formatted using the mutate function to standard mdy
nyc_ps_data <- nyc_ps_data %>% mutate(OCCUR_DATE = mdy(nyc_ps_data$OCCUR_DATE))
#check to see that the format has changed
#class(nyc_ps_data$OCCUR_DATE)
# Formatting the time
#time_data <- nyc_ps_data %>% select(INCIDENT_KEY, OCCUR_TIME)
nyc_ps_data$OCCUR_TIME <- hms(nyc_ps_data$OCCUR_TIME)
#check to see that the time format has changed
#class(nyc_ps_data$OCCUR_TIME)
```

****
### Converting Null values to UNKNOWN for a few columns
```{r,echo=TRUE}
 for(i in 1:nrow(nyc_ps_data)) {
   
   if(nyc_ps_data$LOCATION_DESC[i] == "") {
     nyc_ps_data$LOCATION_DESC[i] <- "UNKNOWN LOC DESC"
   }
   if(nyc_ps_data$PERP_AGE_GROUP[i] == "") {
    nyc_ps_data$PERP_AGE_GROUP[i] <- "UNKNOWN PERP AGE GROUP"
  }
  if(nyc_ps_data$PERP_SEX[i] == "") {
    nyc_ps_data$PERP_SEX[i] <- "UNKNOWN PERP SEX"
  }
  if(nyc_ps_data$PERP_RACE[i] == "") {
    nyc_ps_data$PERP_RACE[i] <- "UNKNOWN PERP RACE"
  }
}
```
### "Factor"izing the categorical variables
```{r,echo=TRUE}
nyc_ps_data$BORO <- as.factor(nyc_ps_data$BORO)
nyc_ps_data$LOCATION_DESC <- as.factor(nyc_ps_data$LOCATION_DESC)
nyc_ps_data$STATISTICAL_MURDER_FLAG <- as.factor(nyc_ps_data$STATISTICAL_MURDER_FLAG)
nyc_ps_data$PERP_AGE_GROUP <- as.factor(nyc_ps_data$PERP_AGE_GROUP)
nyc_ps_data$PERP_SEX <- as.factor(nyc_ps_data$PERP_SEX)
nyc_ps_data$PERP_RACE <- as.factor(nyc_ps_data$PERP_RACE)
nyc_ps_data$VIC_AGE_GROUP <- as.factor(nyc_ps_data$VIC_AGE_GROUP)
nyc_ps_data$VIC_SEX <- as.factor(nyc_ps_data$VIC_SEX)
nyc_ps_data$VIC_RACE <- as.factor(nyc_ps_data$VIC_RACE)
```
### Structure and Summary
```{r,echo=FALSE}
# Structure of the data set
str(nyc_ps_data)
# Sumamry information of the data set
summary(nyc_ps_data)
```
****
### Duplicate INCIDENT Identifcation
``` {r,echo=TRUE}
# Identifying duplicate Incidents and storing in a dataframe
# Duplicate INCIDENTS indicate more than one shooting/victim
incident_vector <- nyc_ps_data$INCIDENT_KEY
incident_df <- data.frame(incident_vector[duplicated(incident_vector)])
incident_df <- distinct(incident_df)
names(incident_df)[1] <- 'INCIDENT_KEY'

dup_incident_ps_data <- nyc_ps_data %>% inner_join(incident_df, by = c("INCIDENT_KEY"))
str(dup_incident_ps_data)

```
****
### Data Summarization
``` {r,echo=TRUE}
# Summarization by Boroughs

boro_totals <- nyc_ps_data %>% group_by(BORO, STATISTICAL_MURDER_FLAG) %>% summarise(totals_by_boro = n())

# Grouping of all incidents by Borough and Year
nyc_data_by_year <- nyc_ps_data %>% mutate(inc_year = year(OCCUR_DATE))  %>% 
  select(c(BORO:VIC_RACE),inc_year)
str(nyc_data_by_year)

# Summarization of all incidents by Borough and Year
nyc_data_summary_by_boro_year <- nyc_data_by_year %>% group_by(BORO,inc_year) %>% 
  summarize(totals_by_boro_year = n())

# Summarization of all incidents by Year
nyc_data_summary_by_year <- nyc_data_summary_by_boro_year %>% group_by(inc_year) %>% 
  summarize(totals_by_year =sum(totals_by_boro_year))

# Summarization of duplicate INCIDENTS by a set of values
dup_boro_totals_by_date <- dup_incident_ps_data %>% group_by(INCIDENT_KEY, BORO, OCCUR_DATE, OCCUR_TIME, LOCATION_DESC, PRECINCT, VIC_SEX, STATISTICAL_MURDER_FLAG) %>% summarise(totals = n()) %>% select(INCIDENT_KEY, BORO, OCCUR_DATE, OCCUR_TIME, LOCATION_DESC, PRECINCT, VIC_SEX, STATISTICAL_MURDER_FLAG,totals)
head(dup_boro_totals_by_date %>% select(INCIDENT_KEY, BORO,STATISTICAL_MURDER_FLAG, totals, everything()))
tail(dup_boro_totals_by_date %>% select(INCIDENT_KEY, BORO,STATISTICAL_MURDER_FLAG, totals, everything()))

# Grouping of duplicates incidents by Borough 
dup_boro_totals <- dup_boro_totals_by_date %>% group_by(BORO, STATISTICAL_MURDER_FLAG) %>% summarise(total_incidents = sum(totals))
head(dup_boro_totals)

# Grouping of duplicate incidents by Borough and Year
dup_boro_totals_by_year <- dup_boro_totals_by_date %>% mutate(dup_year = year(OCCUR_DATE)) %>% select(BORO, dup_year, STATISTICAL_MURDER_FLAG)

# Summarization of duplicate incidents by Borough and Year
dup_boro_summary_by_year <- dup_boro_totals_by_year %>% group_by(BORO, dup_year) %>% summarize(dup_totals_by_year = n())

```
****
### Plots
```{r,echo=TRUE}
# Plot of incident totals by Borough and classification of murder vs injury
boro_totals %>%
  ggplot(aes(fill=STATISTICAL_MURDER_FLAG, x=BORO,y=totals_by_boro)) + 
  geom_bar(position = 'dodge', stat ='identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab("Borough") + ylab("Incidents by Borough")

# Plot of duplicate incident totals by Borough and classification of murder vs injury
dup_boro_totals %>%
  ggplot(aes(fill = STATISTICAL_MURDER_FLAG,x=BORO, y=total_incidents)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab("Borough") + ylab("Duplicate Incidents by Borough")

# Plot of summary of duplicate incident totals by Borough by Year
dup_boro_summary_by_year %>%
  ggplot(aes(fill = BORO,x=dup_year, y=dup_totals_by_year)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab("Year") + ylab("Duplicate Incident totals by Borough")


# Plot of summary of all incident totals by Borough by Year
nyc_data_summary_by_boro_year %>%
  ggplot(aes(fill = BORO,x=inc_year, y=totals_by_boro_year)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab("Year") + ylab("Totals by Borough")

# Plot of summary of duplicate incident totals by Borough by Year >= 2017
dup_boro_summary_by_year %>% filter(dup_year >= 2017) %>% 
  ggplot(aes(fill = BORO,x=dup_year, y=dup_totals_by_year)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab(" Year >= 2017") + ylab("Duplicate Incident totals by Borough")

# Plot of summary of all incident totals by Borough by Year >= 2017
nyc_data_summary_by_boro_year %>% filter(inc_year >= 2017) %>%
  ggplot(aes(fill = BORO,x=inc_year, y=totals_by_boro_year)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab(" Year >= 2017") + ylab("Totals by Borough")

# The plots with Year >= 2017 has been selected and drawn with an underlying bias that the # of incidents per year is increasing after
# it slowly fell
```
****
### Modeling the data
```{r,echo=TRUE}
# Creating a model with total incidents per year 
# totals_by_year = estimate of the intercept + inc_year*estimate
nyc_ps_model <- lm(totals_by_year ~ inc_year, data = nyc_data_summary_by_year)
summary(nyc_ps_model)

# Creating the dataframe for the model 
nyc_data_model_by_year <- nyc_data_summary_by_year %>% mutate(pred_totals = predict(nyc_ps_model))

# Plotting the model against the actuals by Year
# Actual values in blue
# Predictions in red
# totals_by_year = 89177.93 + inc_year*(-43.50)
nyc_data_model_by_year %>%
  ggplot() +
  geom_point(aes(x=factor(inc_year), y=totals_by_year),color = "blue") +
  geom_point(aes(x=factor(inc_year), y=pred_totals),color = "red") +  
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  xlab("Year") + ylab("Total Incidents by Year")
```